### 2013-06-06 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updates to spn intro section

The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \citep{evans_denovo:_2010}. For full reactor
core neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the responses of interest
\citep{slaybaugh_acceleration_2011}. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

@@

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment with both variational and
asymptotic derivations performed as a means of verification. In
addition, these equations have been more rigorously studied as
solution methods to MOX fuel problems and have been shown to provide
accurate solutions. With this mathematical literature to provide a
solid numerical footing for the method, we look at its application to
today's challenge problems in full reactor core transport. The
reduction in numerical complexity of current full core deterministic
solution methods using the $S_N$ approximation could mean significant
savings in both compute time and memory required. In addition, the
characteristics of the solution to the transport problem for a steady
state reactor core permit diffusion theory to be used; a staple of the
nuclear industry since its inception. Therefore, if diffusion theory
is applicable, then finer grained solutions that capture more of the
physics contained in the transport equation should be possible with
the $SP_N$ method. In doing so, we also expect from the literature to
obtain computed responses on the order of accuracy we would expect
from an appropriately discretized $S_N$ method at a fraction of the
cost.

To further motivate moving in this direction, recent developments in
the Exnihilo neutronics package at Oak Ridge National Laboratory have
permitted generation of the $SP_N$ system of equations for detailed
full core reactor models \citep{evans_simpli_2013}. By fully forming
these equations and formulating them as a linear algebra problem
instead of using the explicit iterative methods of the past, we now
have access to all of the modern advancements in computational linear
algebra including Krylov solvers for asymmetric systems and
preconditioning methods such as algebraic multigrid. This leads us to
then explore the applicability of our work in discrete Monte Carlo
methods for linear systems as a possible solution method for the
$SP_N$ equations. If formulated correctly, we hypothesize that
significant improvement in usage of computational resources may be
observed compared to modern solution techniques such as those
suggested due to the form of the matrices generated by the
discretization. In addition, solving the $SP_N$ equations in this way
also breaks away from the $S_N$ forms of parallelism where spatial
parallelism is achieved by an efficient parallel sweep, angular
efficiency achieved by pipe-lining, and energy parallelism achieved by
decoupling the groups. With the $SP_N$ equations as a full matrix
system, we now can parallelize the problem as prescribed by the linear
solver, which may be significantly more scalable than current $S_N$
transport practices.

In this chapter, we derive the $SP_N$ equations, closely following the
work of Evans\footnote{The $SP_N$ derivations in this chapter are
  heavily based on those presented by Evans in
  \citep{evans_simpli_2013}}, in order to gain full understanding of
the underlying system and its potential behavior in a Monte Carlo
context. From the $P_N$ equations, we apply a set of approximations to
yield the multi-dimensional, multi-group $SP_N$ equations for fixed
source problems. Using the fully-formed matrix for the transport
problem, we explore solutions to the $SP_N$ equations using Monte
Carlo Synthetic Acceleration using a light water reactor fuel assembly
criticality calculation as the driving problem.

%%---------------------------------------------------------------------------%%
\section{The Neutron Transport Equation}
\label{sec:transport_eq}
As a starting point we define the time-independent neutron transport
equation \citep{lewis_computational_1993}:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
  q(\vec{r},\hat{\Omega},E)\:,
  \label{eq:general_transport}
\end{multline}
with the variables defined as:
\begin{itemize}
\item $\vec{r}$ - neutron spatial position
\item $\hat{\Omega}$ - neutron streaming direction with radial
  component $\mu$ and azimuthal component $\omega$
\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
  scattering
\item $E$ - neutron energy
\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
\item $\sigma(\vec{r},E)$ - total interaction cross section
\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
  of scattering from direction $\hat{\Omega}'$ into an angular domain
  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
  $E'$ to an energy domain $dE'$ about energy $E$
\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
\end{itemize}
For this work, it is sufficient to formulate
Eq~(\ref{eq:general_transport}) in 1-dimensional Cartesian geometry:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
  \frac{q(x,E)}{4 \pi}\:,
  \label{eq:cart_1d_transport}
\end{multline}
where the angular component of the solution is no longer dependent on
the azimuthal direction of travel and an isotropic source of neutrons
is assumed. In addition, for fission systems the eigenvalue form of
the transport equation is:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
  \label{eq:eigenvalue_transport}
\end{multline}
with the additional variables defined as
\begin{itemize}
\item $k$ - multiplication factor
\item $\chi(E)$ - fission neutron energy spectrum
\item $\nu$ - average number of neutrons per fission
\item $\sigma_f(r,E')$ - fission cross section\:.
\end{itemize}
In 1-dimensional Cartesian geometry,
Eq~(\ref{eq:eigenvalue_transport}) becomes:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
  dE' + \frac{q(x,E)}{4 \pi}\:.
  \label{eq:cart_1d_eigenvalue}
\end{multline}

>>>>

The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \citep{evans_denovo:_2010}. For fission
reactor neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the responses of interest
\citep{slaybaugh_acceleration_2011}. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

@@

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment with both variational and
asymptotic derivations performed as a means of verification. In
addition, these equations have been more rigorously studied as
solution methods to MOX fuel problems and have been shown to provide
accurate solutions. With this mathematical literature to provide a
solid numerical footing for the method, we look at its application to
today's challenge problems in neutron transport for fission reactor
analysis. The reduction in numerical complexity of current
deterministic solution methods using the $S_N$ approximation could
mean significant savings in both compute time and memory required. In
addition, the characteristics of the solution to the transport problem
for a steady state reactor core permit diffusion theory to be used; a
staple of the nuclear industry since its inception. Therefore, if
diffusion theory is applicable, then finer grained solutions that
capture more of the physics contained in the transport equation should
be possible with the $SP_N$ method. In doing so, we also expect from
the literature to obtain computed responses on the order of accuracy
we would expect from an appropriately discretized $S_N$ method at a
fraction of the cost.

To further motivate moving in this direction, recent developments in
the Exnihilo neutronics package at Oak Ridge National Laboratory have
permitted generation of the $SP_N$ system of equations for detailed
neutronics models \citep{evans_simpli_2013}. By fully forming these
equations and formulating them as a linear algebra problem instead of
using the explicit iterative methods of the past, we now have access
to all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and preconditioning
methods such as algebraic multigrid. This leads us to then explore the
applicability of our work in discrete Monte Carlo methods for linear
systems as a possible solution method for the $SP_N$ equations. If
formulated correctly, we hypothesize that significant improvement in
usage of computational resources may be observed compared to modern
solution techniques such as those suggested due to the form of the
matrices generated by the discretization. In addition, solving the
$SP_N$ equations in this way also breaks away from the $S_N$ forms of
parallelism where spatial parallelism is achieved by an efficient
parallel sweep, angular efficiency achieved by pipe-lining, and energy
parallelism achieved by decoupling the groups. With the $SP_N$
equations as a full matrix system, we now can parallelize the problem
as prescribed by the linear solver, which may be significantly more
scalable than current $S_N$ transport practices.

In this chapter, we derive the $SP_N$ equations, closely following the
work of Evans\footnote{The $SP_N$ derivations in this chapter are
  heavily based on those presented by Evans in
  \citep{evans_simpli_2013}}, in order to gain full understanding of
the underlying system and its potential behavior in a Monte Carlo
context. From the $P_N$ equations, we apply a set of approximations to
yield the $SP_N$ equations for fixed source and criticality
problems. Using the fully-formed linear operator for the transport
problem, we explore solutions to the $SP_N$ equations using Monte
Carlo Synthetic Acceleration using a light water reactor fuel assembly
criticality calculation as the driving problem. In order to
effectively solve this problem, a suite of preconditioners is
developed and studied within the context of MCSA. Using these
preconditioners, MCSA solutions for the fuel assembly problem are
verified by comparing against production Krylov solvers and finally
performance is analyzed through comparison with those same Krylov
solvers.

%%---------------------------------------------------------------------------%%
\section{The Neutron Transport Equation}
\label{sec:transport_eq}
As a starting point for the $SP_N$ equations we define the
time-independent neutron transport equation
\citep{lewis_computational_1993}:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
  q(\vec{r},\hat{\Omega},E)\:,
  \label{eq:general_transport}
\end{multline}
with the variables defined as:
\begin{itemize}
\item $\vec{r}$ - neutron spatial position
\item $\hat{\Omega}$ - neutron streaming direction with radial
  component $\mu$ and azimuthal component $\omega$
\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
  scattering
\item $E$ - neutron energy
\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
\item $\sigma(\vec{r},E)$ - total interaction cross section
\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
  of scattering from direction $\hat{\Omega}'$ into an angular domain
  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
  $E'$ to an energy domain $dE'$ about energy $E$
\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
\end{itemize}
For this work, it is sufficient to formulate
Eq~(\ref{eq:general_transport}) in 1-dimensional Cartesian geometry:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
  \frac{q(x,E)}{4 \pi}\:,
  \label{eq:cart_1d_transport}
\end{multline}
where the angular component of the solution is no longer dependent on
the azimuthal direction of travel and an isotropic source of neutrons
is assumed. In addition, for fission systems the eigenvalue form of
the transport equation is:
\begin{multline}
  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
  \label{eq:eigenvalue_transport}
\end{multline}
with the additional variables defined as
\begin{itemize}
\item $k$ - multiplication factor
\item $\chi(E)$ - fission neutron energy spectrum
\item $\nu$ - average number of neutrons per fission
\item $\sigma_f(r,E')$ - fission cross section\:.
\end{itemize}
In 1-dimensional Cartesian geometry,
Eq~(\ref{eq:eigenvalue_transport}) becomes:
\begin{multline}
  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
  dE' + \frac{q(x,E)}{4 \pi}\:.
  \label{eq:cart_1d_eigenvalue}
\end{multline}
